---
title: "L'ORA - WG1 Working example Spillover (so)"
author: "Alfredo Acosta SVA" 
date: "`r Sys.Date()`" 
output: 
  html_document:
    toc: yes
    toc_float: true
    theme: cosmo
    fig_caption: yes
    number_sections: yes
    code_folding: show
  html_notebook: 
    toc: yes
---
# Working example SpillOver
-The code use several datasets EMPRES-I (FAO), ECDC Surveillance Atlas ECDC, Global Health observatory (WHO). Adagio.
-Please feel free to analyse the code and run from your terminal (there is a folder with the code and data on: DG1 Disease occurrence > datasources > Spill-over > sources > spillover.Rmd)

# Libraries
```{r}
# Spill over analysis
# Update: 17/01/2024
library(tidyverse)
library(lubridate)
library(stringr)
library(readr)
library(data.table)
library(plotly)
library(DT)
```
# Working directory
```{r}
setwd("C:/Users/alfredo.acosta/SVA/LiRA_consortium - Documents/WG1 Disease occurrence/datasources/Spill-over/sources")

```

# Loading ECDC Surveillance atlas
```{r}
# West Nile 
soe <- read.csv(file="ECDC_surveillance_data_West_Nile_virus_infection.csv")
# Rabies
soe1 <- read.csv(file="ECDC_surveillance_data_Rabies.csv")
# Crimean congo
soe2 <- read.csv(file="ECDC_surveillance_data_Crimean-Congo_haemorrhagic_fever.csv")
# Rift Valley
soe3 <- read.csv(file="ECDC_surveillance_data_Rift_valley_fever.csv")
# Influenza
# soe4 <- read.csv(file="ECDC_surveillance_data_Influenza.csv")

soe <- rbind(soe, soe1, soe2, soe3)
rm(soe1, soe2, soe3)

soe1 <- soe %>% 
    group_by(HealthTopic, RegionName,Time) %>%
    summarise(cases=sum(NumValue)) %>% 
  filter(cases != 0)

colnames(soe1) <- c("disease", "country", "time", "cases")

```
# Descriptive
```{r}
# Number of human cases
soe %>% 
  filter(!str_detect(RegionName, "EU")) %>% 
    group_by(HealthTopic) %>%
  summarise(cases=sum(NumValue)) %>% 
  filter(cases != 0)

# Number of countries
soe %>% 
  filter(!str_detect(RegionName, "EU")) %>% 
    group_by(HealthTopic) %>%
  summarise(countries=length(unique(RegionName)))

table(soe$Time)
table(soe$Time, soe$HealthTopic)

```
# Loading Adagio database
```{r}
# Reading ADAGIO database
adg <- read.csv(file= "Outbreaks and cases with labels.csv")
str(adg)

adg %>% 
  filter(Disease == "West Nile Fever") %>% 
  group_by(Country, FunctionalGroup,Species, Year) %>% 
  summarize(cases=sum(SumCases))

adg %>% 
  filter(Disease == "West Nile Fever") %>% 
  # group_by(Country, FunctionalGroup,Species, Year) %>% 
  filter(Species != "Domesticated Equids") %>%
  summarize(cases=sum(SumCases, na.rm = TRUE))

# filtering cases by country without equids
adg %>% 
  filter(Disease == "West Nile Fever") %>% 
  filter(Species != "Domesticated Equids") %>%
  group_by(Country) %>% 
  summarize(cases=sum(SumCases))


```

```{r}
# West Nile animals cases to match human cases
adg2  <- adg %>% 
  filter(Species != "Domesticated Equids") %>%
  mutate(Disease = gsub("West Nile Fever", "West Nile virus infection", Disease)) %>% 
  # filter(Disease == "West Nile virus infection") %>% 
  group_by(Country, Disease, Year) %>% 
  summarize(cases=sum(SumCases))

adg2

sum(adg2$cases, na.rm = TRUE) #1720-915  #total 814 withouth equids

```
# Loading Human population
```{r}
hpop <- read_csv("API_SP.POP.TOTL_DS2_en_csv_v2_6508519.csv", skip = 3)
hpop$`Country Name` <- gsub("Kosovo*","Kosovo", hpop$`Country Name`)

colnames(hpop)
hpop2 <- hpop[,c(1,48:66)]
hpop2 <- melt(hpop2, id = 1)
colnames(hpop2)
```

# Matrix with human population over the years
```{r}
# Create matrix
db <-expand.grid(unique(hpop$`Country Name`), c(2003:2023), unique(soe1$disease))

colnames(db) <- c("country", "time", "disease")

# First index to pass the animal cases db & adg
index <- match(paste(db$country, db$time, db$disease), paste(adg2$Country, adg2$Year, adg2$Disease))
db$animal_c <- adg2$cases[index]

# Second index to pass the human cases db & soe1
index <- match(paste(db$country, db$time, db$disease), paste(soe1$country, soe1$time, soe1$disease))
db$human_c <- soe1$cases[index]

unique(db$disease)
unique(soe1$disease)

# third index to pass the human pop
index <- match(paste(db$country, db$time), paste(hpop2$`Country Name`, hpop2$variable))
db$human_pop <- hpop2$value[index]

db %>% 
  group_by(disease) %>%
  summarise(casesAnimal=sum(animal_c, na.rm = TRUE),
            casesHuman=sum(human_c, na.rm = TRUE))
```

## Human cases / 100.000 habitants
```{r}
db$human_cases_hab <- db$human_c/db$human_pop*100000
```

# Reading the UN regions
```{r}
unregions <- read.csv("all.csv")
unregions$name[unregions$name == "United States of America"] <- "United States"
unregions$name[unregions$name == "United Kingdom of Great Britain and Northern Ireland"] <- "United Kingdom"

str(unregions)
str(db)
db$country <- as.character(db$country)
db <- data.frame(db)
db$unr <- unregions$sub.region[match(db$country, unregions$name)]

```
# Update H1N1 - RVF - JE FAO report 
```{r}
so <- read.csv(file="Latest Reported Events.csv") #spill-over: so
so1 <- read.csv(file="Rift Valley fever (all).csv") #spill-over: so

so$Serotype <- NULL
so1$Location <- NULL

colnames(so)[3] <- "lat"
colnames(so)[4] <- "lon"
colnames(so1)

so1 <- so1[, c(1:9,13,17,15,16,14)]
so <- rbind(so, so1)

so$Country[so$Country == "U.K. of Great Britain and Northern Ireland"] <- "United Kingdom"
so$Country[so$Country == "United States of America"] <- "United States"


so$year <- year(ymd(substr(so$report.date, 1,10)) )

# Region distribution for cases (lira interest)
fao <- so %>% 
  filter(!is.na(Humans.Affected)) %>% 
  filter(Disease == "Influenza - Avian" |
           Disease == "Rabies" | 
           Disease == "Rift Valley fever") %>% 
  group_by(country=Country, time=year, disease=Disease) %>% 
  summarise(human_c=sum(Humans.Affected))

fao_animal <- so %>% 
  filter(is.na(Humans.Affected)) %>% 
  filter(Disease == "Influenza - Avian" |
           Disease == "Rabies" | 
           Disease == "Rift Valley fever") %>% 
  group_by(country=Country, time=year, disease=Disease) %>% 
  summarise(animal_c=n())

fao$animal_c <- fao_animal$animal_c[match(paste(fao$country, fao$time, fao$disease), paste(fao_animal$country, fao_animal$time, fao_animal$disease))]


fao <- fao[,c(1,2,3,5,4)]


colnames(db)
colnames(fao)

# Completing with population
fao$human_pop <- hpop2$value[match(paste(fao$country, fao$time-2), paste(hpop2$`Country Name`, hpop2$variable))]
fao$human_cases_hab <- fao$human_c/fao$human_pop*100000

fao$unr <- unregions$sub.region[match(fao$country, unregions$name)]

# Japanese encephalitis
# https://www.who.int/data/gho/data/indicators/indicator-details/GHO/japanese-encephalitis---number-of-reported-cases
so2 <- read.csv(file="FAO_JE.csv") #spill-over: so

str(so2)

so3 <- so2 %>% 
  mutate(disease=gsub("Japanese encephalitis - number of reported cases", "Japanese encephalitis", Indicator)) %>%
  mutate(animal_c= "0") %>% 
  filter(Value != 0) %>% 
  group_by(country=Location, time=Period, disease, animal_c) %>% 
  summarise(human_c=sum(Value))

# Completing with population
so3$human_pop <- hpop2$value[match(paste(so3$country, so3$time), paste(hpop2$`Country Name`, hpop2$variable))]

so3$human_cases_hab <- so3$human_c/so3$human_pop*100000

so3$unr <- unregions$sub.region[match(so3$country, unregions$name)]

colnames(db)
colnames(so3)

db <- rbind(db, fao, so3)

```
# General view on rate x 100.000 hab
```{r}
db %>% 
  filter(time > 2003) %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(time, human_cases_hab, group=country))+
  geom_point()

db %>% 
  filter(time > 2003) %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(time, human_cases_hab, group=country))+
  geom_point()+
  scale_y_log10()+
  facet_wrap(vars(disease), ncol = 1)
  
```
## West Nile Virus
```{r}
db %>% 
  filter(time > 2003) %>% 
  filter(disease == "West Nile virus infection") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, color=country))+
  geom_boxplot()+
  scale_x_log10()+
  facet_wrap(vars(country), ncol = 1)+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("WNV human cases / 100.000 habitants")+
  ylab(NULL) + 
  guides(y="none")

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "West Nile virus infection") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, time, group=time))+
  geom_boxplot()+
  scale_x_log10()+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("WNV human cases / 100.000 habitants (boxplot grouped by year)")+
  ylab(NULL)

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "West Nile virus infection") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(time, human_cases_hab, color=country))+
  geom_point()+
  geom_line(size=1)+
  scale_y_log10()+
  facet_wrap(vars(country), ncol = 3)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("WNF human cases / 100.000 habitants (by year-country)")
```

## Rabies
```{r}
db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Rabies") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, time, group=time))+
  geom_boxplot()+
  scale_x_log10()+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("Rb human cases / 100.000 habitants (boxplot grouped by year)")+
  ylab(NULL)

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Rabies") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, color=country))+
  geom_boxplot()+
  scale_x_log10()+
  facet_wrap(vars(country), ncol = 1)+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("Rb human cases / 100.000 habitants")+
  ylab(NULL) + 
  guides(y="none")

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Rabies") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(time, human_cases_hab, color=country))+
  geom_point()+
  geom_line(size=1)+
  scale_y_log10()+
  facet_wrap(vars(country), ncol = 3)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("Rb human cases / 100.000 habitants (by year-country)")


```

## Rift valley fever
```{r}
db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Rift valley fever") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, time, group=time))+
  geom_boxplot()+
  scale_x_log10()+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("RVF human cases / 100.000 habitants (boxplot grouped by year)")+
  ylab(NULL)

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Rift valley fever") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, color=country))+
  geom_boxplot()+
  scale_x_log10()+
  facet_wrap(vars(country), ncol = 1)+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("RVF human cases / 100.000 habitants")+
  ylab(NULL) + 
  guides(y="none")

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Rift valley fever") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(time, human_cases_hab, color=country))+
  geom_point()+
  geom_line(size=1)+
  scale_y_log10()+
  facet_wrap(vars(country), ncol = 3)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("RVF human cases / 100.000 habitants (by year-country)")

```

## Crimean-Congo haemorrhagic fever
```{r}
db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Crimean-Congo haemorrhagic fever") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, time, group=time))+
  geom_boxplot()+
  scale_x_log10()+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("CCHF human cases / 100.000 habitants (boxplot grouped by year)")+
  ylab(NULL)

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Crimean-Congo haemorrhagic fever") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, color=country))+
  geom_boxplot()+
  scale_x_log10()+
  facet_wrap(vars(country), ncol = 1)+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("CCHF human cases / 100.000 habitants")+
  ylab(NULL) + 
  guides(y="none")

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Crimean-Congo haemorrhagic fever") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(time, human_cases_hab, color=country))+
  geom_point()+
  geom_line(size=1)+
  scale_y_log10()+
  facet_wrap(vars(country), ncol = 3)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("CCHF human cases / 100.000 habitants (by year-country)")

```
## Influenza - Avian
```{r}
db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Influenza - Avian") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, time, group=time))+
  geom_boxplot()+
  scale_x_log10()+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("CCHF human cases / 100.000 habitants (boxplot grouped by year)")+
  ylab(NULL)

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Influenza - Avian") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, color=country))+
  geom_boxplot()+
  scale_x_log10()+
  facet_wrap(vars(country), ncol = 1)+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("CCHF human cases / 100.000 habitants")+
  ylab(NULL) + 
  guides(y="none")

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Influenza - Avian") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(time, human_cases_hab, color=country))+
  geom_point()+
  geom_line(size=1)+
  scale_y_log10()+
  facet_wrap(vars(country), ncol = 3)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("CCHF human cases / 100.000 habitants (by year-country)")

```

## Japanese encephalitis
```{r}
db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Japanese encephalitis") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, time, group=time))+
  geom_boxplot()+
  scale_x_log10()+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("CCHF human cases / 100.000 habitants (boxplot grouped by year)")+
  ylab(NULL)

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Japanese encephalitis") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, color=country))+
  geom_boxplot()+
  scale_x_log10()+
  facet_wrap(vars(country), ncol = 1)+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("CCHF human cases / 100.000 habitants")+
  ylab(NULL) + 
  guides(y="none")

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Japanese encephalitis") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(time, human_cases_hab, color=country))+
  geom_point()+
  geom_line(size=1)+
  scale_y_log10()+
  facet_wrap(vars(country), ncol = 3)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("CCHF human cases / 100.000 habitants (by year-country)")

```
# Ordering number of cases to check patterns 
```{r}
wnf <- db %>% 
  group_by(country, disease) %>% 
  filter(time > 2003) %>% 
  filter(!is.na(human_cases_hab)) %>% 
  summarise(mean_cases=mean(human_cases_hab)) %>% 
  # mutate(level=cut(mean_cases, breaks = 3, labels = c(1,2,3))) %>% 
  mutate(level=cut(mean_cases, breaks = 3)) %>% 
  arrange(desc(level))

dbw <- db %>% 
  filter(time > 2003) %>% 
  filter(disease == "West Nile virus infection") %>% 
  filter(!is.na(human_cases_hab))

dbw$order <- wnf$mean_cases[match(dbw$country, wnf$country)]
dbw$country <- reorder(dbw$country, dbw$order)
 
dbw %>% 
  filter(disease == "West Nile virus infection") %>% 
  ggplot(aes(time, human_cases_hab, color=country))+
  geom_point()+
  geom_line(size=0.1)+
  scale_y_log10()+
  facet_wrap(vars(country), nrow = 3)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("WNF human cases / 100.000 habitants (by year-country) ordered")
```

## Rabies ordering
```{r}

wnf <- db %>% 
  group_by(country) %>% 
  filter(time > 2003) %>% 
  filter(disease == "Rabies") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  summarise(mean_cases=mean(human_cases_hab)) %>% 
  # mutate(level=cut(mean_cases, breaks = 3, labels = c(1,2,3))) %>% 
  mutate(level=cut(mean_cases, breaks = 3)) %>% 
arrange(desc(level))


dbw <- db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Rabies") %>% 
  filter(!is.na(human_cases_hab))

dbw$order <- wnf$mean_cases[match(dbw$country, wnf$country)]

dbw$country <- reorder(dbw$country, dbw$order)
 

  ggplot(dbw,aes(time, human_cases_hab, color=country))+
  geom_point()+
  geom_line(size=0.1)+
  scale_y_log10()+
  facet_wrap(vars(country), nrow = 3)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("RB human cases / 100.000 habitants country ordered by the mean rate")

```


# General view of diseases spread
```{r}
db %>% 
  filter(time > 2003) %>% 
  # filter(disease == "Rabies") %>% 
  # filter(unr != "Western Asia") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, time, group=time, color=disease))+
  geom_boxplot()+
  scale_x_log10()+
  facet_wrap(vars(disease), ncol = 1)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("human cases/100.000 habitants (log scaled) country rates")

db %>% 
  filter(time > 2003 & !is.na(human_cases_hab & !is.na(disease))) %>% 
  # filter(disease == "Rabies") %>% 
  # filter(!is.na(human_cases_hab)) %>% 
  # filter(unr != "Western Asia") %>%
  ggplot(aes(human_cases_hab, unr, group=unr, color=disease))+
  geom_boxplot()+
  scale_x_log10()+
  facet_wrap(vars(disease), scales="free", ncol = 1)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("human cases/100.000 habitants (log scaled) country rates ")
```
# Calculating cut points and levels by disease
```{r}
level <- db %>% 
  filter(time > 2003) %>%
  # filter(!is.na(unr)) %>% 
  filter(!is.na(human_cases_hab)) %>% 
  group_by(disease) %>% 
  mutate(name=cut(human_cases_hab, breaks = 3, labels = c("low","medium","high"))) %>% 
  mutate(ranges=cut(human_cases_hab, breaks = 3)) %>% 
  select(-unr, -animal_c)

level2 <- level %>%  
  group_by(disease,name,ranges) %>% 
  dplyr::summarize(mean_point=median(human_cases_hab),
                   human_cases_hab_unr=(sum(human_c)/sum(human_pop*100000)),
                   mean2=mean(human_cases_hab_unr))

# other option please consider and improve not yet usable because the breaks is static and we need to be disease specific
to_break <- db %>% 
  filter(time > 2003) %>%
  filter(!is.na(unr)) %>% 
  filter(!is.na(human_cases_hab))
breaks <- quantile(to_break$human_cases_hab, c(0,0.25,0.75,1))
breaks

level <- to_break %>% 
  group_by(disease,) %>% 
  mutate(name=cut(human_cases_hab, breaks = breaks, labels = c("low","medium","high"))) %>% 
  mutate(ranges=cut(human_cases_hab, breaks = breaks, right = FALSE)) %>% 
  mutate(name2=cut(human_cases_hab, breaks = 3, labels = c("low","medium","high"))) %>% 
  mutate(ranges2=cut(human_cases_hab, breaks = 3, right = FALSE)) %>% 
  select(-unr, -animal_c)

```

# Using UN regions
## West Nile Virus
```{r}
db %>% 
  filter(time > 2003) %>% 
  filter(disease == "West Nile virus infection") %>% 
  filter(!is.na(human_cases_hab)) %>% 
   filter(!is.na(unr)) %>% 
  ggplot(aes(human_cases_hab, color=unr))+
  geom_boxplot()+
  scale_x_log10()+
  facet_wrap(vars(unr), ncol = 1)+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("WNV human cases / 100.000 habitants")+
  ylab(NULL) + 
  guides(y="none")

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "West Nile virus infection") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  filter(!is.na(unr)) %>% 
  ggplot(aes(human_cases_hab, time, group=time))+
  geom_boxplot()+
  scale_x_log10()+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("WNV human cases / 100.000 habitants (boxplot grouped by year)")+
  ylab(NULL)

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "West Nile virus infection") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  filter(!is.na(unr)) %>% 
  ggplot(aes(time, human_cases_hab, color=unr))+
  geom_point()+
  geom_line(size=1)+
  scale_y_log10()+
  facet_wrap(vars(unr), ncol = 3)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("WNF human cases / 100.000 habitants (by year-country)")
```

## Rabies
```{r}
db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Rabies") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, time, group=time))+
  geom_boxplot()+
  scale_x_log10()+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("Rb human cases / 100.000 habitants (boxplot grouped by year)")+
  ylab(NULL)

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Rabies") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, color=unr))+
  geom_boxplot()+
  scale_x_log10()+
  facet_wrap(vars(unr), ncol = 1)+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("Rb human cases / 100.000 habitants")+
  ylab(NULL) + 
  guides(y="none")

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Rabies") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(time, human_cases_hab, color=unr))+
  geom_point()+
  geom_line(size=1)+
  scale_y_log10()+
  facet_wrap(vars(unr), ncol = 3)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("Rb human cases / 100.000 habitants (by year-country)")


```

## Rift valley fever
```{r}
db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Rift valley fever") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, time, group=time))+
  geom_boxplot()+
  scale_x_log10()+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("RVF human cases / 100.000 habitants (boxplot grouped by year)")+
  ylab(NULL)

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Rift valley fever") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, color=unr))+
  geom_boxplot()+
  scale_x_log10()+
  facet_wrap(vars(unr), ncol = 1)+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("RVF human cases / 100.000 habitants")+
  ylab(NULL) + 
  guides(y="none")

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Rift valley fever") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(time, human_cases_hab, color=unr))+
  geom_point()+
  geom_line(size=1)+
  scale_y_log10()+
  facet_wrap(vars(unr), ncol = 3)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("RVF human cases / 100.000 habitants (by year-country)")

```

## Crimean-Congo haemorrhagic fever
```{r}
db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Crimean-Congo haemorrhagic fever") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, time, group=time))+
  geom_boxplot()+
  scale_x_log10()+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("CCHF human cases / 100.000 habitants (boxplot grouped by year)")+
  ylab(NULL)

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Crimean-Congo haemorrhagic fever") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, color=unr))+
  geom_boxplot()+
  scale_x_log10()+
  facet_wrap(vars(unr), ncol = 1)+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("CCHF human cases / 100.000 habitants")+
  ylab(NULL) + 
  guides(y="none")

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Crimean-Congo haemorrhagic fever") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(time, human_cases_hab, color=unr))+
  geom_point()+
  geom_line(size=1)+
  scale_y_log10()+
  facet_wrap(vars(unr), ncol = 3)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("CCHF human cases / 100.000 habitants (by year-unr)")

```

## Influenza - Avian
```{r}
db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Influenza - Avian") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, time, group=time))+
  geom_boxplot()+
  scale_x_log10()+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("CCHF human cases / 100.000 habitants (boxplot grouped by year)")+
  ylab(NULL)

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Influenza - Avian") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, color=unr))+
  geom_boxplot()+
  scale_x_log10()+
  facet_wrap(vars(unr), ncol = 1)+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("CCHF human cases / 100.000 habitants")+
  ylab(NULL) + 
  guides(y="none")

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Influenza - Avian") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(time, human_cases_hab, color=unr))+
  geom_point()+
  geom_line(size=1)+
  scale_y_log10()+
  facet_wrap(vars(unr), ncol = 3)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("CCHF human cases / 100.000 habitants (by year-unr)")

```
## Japanese encephalitis
```{r}
db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Japanese encephalitis") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, time, group=time))+
  geom_boxplot()+
  scale_x_log10()+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("CCHF human cases / 100.000 habitants (boxplot grouped by year)")+
  ylab(NULL)

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Japanese encephalitis") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, color=unr))+
  geom_boxplot()+
  scale_x_log10()+
  facet_wrap(vars(unr), ncol = 1)+
  theme_minimal() +
  theme(legend.position="none")+
  xlab("CCHF human cases / 100.000 habitants")+
  ylab(NULL) + 
  guides(y="none")

db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Japanese encephalitis") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(time, human_cases_hab, color=unr))+
  geom_point()+
  geom_line(size=1)+
  scale_y_log10()+
  facet_wrap(vars(unr), ncol = 3)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("CCHF human cases / 100.000 habitants (by year-unr)")

```
# Ordering to see patterns
## WNV
```{r}
wnf <- db %>% 
  group_by(unr, disease) %>% 
  filter(time > 2003) %>% 
  filter(!is.na(human_cases_hab)) %>% 
  summarise(mean_cases=mean(human_cases_hab)) %>% 
  # mutate(level=cut(mean_cases, breaks = 3, labels = c(1,2,3))) %>% 
  mutate(level=cut(mean_cases, breaks = 3)) %>% 
  arrange(desc(level))

dbw <- db %>% 
  filter(time > 2003) %>% 
  filter(disease == "West Nile virus infection") %>% 
  filter(!is.na(human_cases_hab))

dbw$order <- wnf$mean_cases[match(dbw$unr, wnf$unr)]
dbw$unr <- reorder(dbw$unr, dbw$order)
 
dbw %>% 
  filter(disease == "West Nile virus infection") %>% 
  ggplot(aes(time, human_cases_hab, color=unr))+
  geom_point()+
  geom_line(size=0.1)+
  scale_y_log10()+
  facet_wrap(vars(unr), nrow = 3)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("WNF human cases / 100.000 habitants (by year-UNregion) ordered")
```

## Rabies
```{r}

wnf <- db %>% 
  group_by(unr) %>% 
  filter(time > 2003) %>% 
  filter(disease == "Rabies") %>% 
  filter(!is.na(human_cases_hab)) %>% 
  summarise(mean_cases=mean(human_cases_hab)) %>% 
  # mutate(level=cut(mean_cases, breaks = 3, labels = c(1,2,3))) %>% 
  mutate(level=cut(mean_cases, breaks = 3)) %>% 
arrange(desc(level))


dbw <- db %>% 
  filter(time > 2003) %>% 
  filter(disease == "Rabies") %>% 
  filter(!is.na(human_cases_hab))

dbw$order <- wnf$mean_cases[match(dbw$unr, wnf$unr)]

dbw$unr <- reorder(dbw$unr, dbw$order)
 

  ggplot(dbw,aes(time, human_cases_hab, color=unr))+
  geom_point()+
  geom_line(size=0.1)+
  scale_y_log10()+
  facet_wrap(vars(unr), nrow = 3)+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("RB human cases / 100.000 habitants UNregion ordered by the mean rate")

  

```


# General view of the rate by UNRegions
```{r}
UN_db <- db %>% 
  group_by(disease, unr, time) %>% 
  filter(time > 2003) %>% 
  filter(unr != "Western Asia") %>% 
  filter(!is.na(human_cases_hab)) %>%
  summarise(human_cases_hab_unr=sum(human_c)/sum(human_pop)*100000)
datatable(UN_db)

ggplotly(UN_db %>% 
  filter(time > 2003) %>% 
  # filter(disease == "Rabies") %>% 
  filter(!is.na(human_cases_hab_unr)) %>% 
  ggplot(aes(human_cases_hab_unr, unr, group=unr, color=disease))+
  geom_boxplot()+
  scale_x_log10()+
  facet_wrap(vars(disease), scales = "free", ncol=1) +
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("human cases / 100.000 habitants (boxplots of UNRegion rates) log scaled "))
```

# Calculating cut points and levels by disease and UNregions
```{r}

UNRegion_db <- db %>% 
  group_by(disease, unr) %>% 
  filter(time > 2003) %>% 
  filter(unr != "Western Asia") %>% 
  filter(!is.na(human_cases_hab)) %>%
  summarise(human_cases_hab_unr=(sum(human_c)/sum(human_pop)*100000)) %>% 
  mutate(name=cut(human_cases_hab_unr, breaks = 3, labels = c("low","medium","high"))) %>% 
  mutate(range=cut(human_cases_hab_unr, breaks = 3))
UNRegion_db


UNRegion_db <- db %>% 
  group_by(disease) %>% 
  filter(time > 2003) %>% 
  filter(unr != "Western Asia") %>% 
  filter(!is.na(human_cases_hab)) %>%
  summarise(human_cases_hab_unr=(sum(human_c)/sum(human_pop)*100000)) %>% 
  mutate(name=cut(human_cases_hab_unr, breaks = 3, labels = c("low","medium","high"))) %>% 
  mutate(range=cut(human_cases_hab_unr, breaks = 3))
UNRegion_db

```

# General cut points by disease
```{r}
db %>% 
  filter(time > 2003) %>% 
  filter(!is.na(human_cases_hab)) %>% 
  ggplot(aes(human_cases_hab, color=disease, group=unr))+
  geom_boxplot()+
  scale_x_log10()+
  facet_wrap(vars(disease, unr))+
  theme_minimal() +
  theme(legend.position="none") +
  ylab(NULL)+
  xlab("human cases / 100.000 habitants (agregated by UNRegions and years)")

```

# Lira table - General cut points by disease values
```{r}
# lira_table_sp <- db %>%
#   filter(time > 2003) %>%
#   # filter(unr != "Western Asia") %>% 
#   filter(!is.na(human_cases_hab)) %>%
#   group_by(disease, UNregions=unr) %>%
#   summarise(
#     human_cases=sum(human_c),
#     human_pop=sum(human_pop),
#     # animal_cases=sum(animal_c),
#     human_cases_hab_UNR=(sum(human_c)/sum(human_pop)*100000),
#     CI05 = quantile(human_cases_hab, probs = 0.05, na.rm = TRUE),
#     CI95 = quantile(human_cases_hab, probs = 0.95, na.rm = TRUE)) %>%
#   mutate(level=cut(human_cases_hab_UNR, breaks = 3, labels = c("low","medium","high"))) %>% 
#   arrange(disease, desc(human_cases_hab_UNR), UNregions)
# 
# lira_table_sp
# write.csv(lira_table_sp, file="lira_table_sp.csv")

```

```{r}
lira_table_sp <- db %>%
  filter(time > 2003) %>%
  # filter(unr != "Western Asia") %>% 
  filter(!is.na(human_cases_hab)) %>%
  group_by(disease, UNregions=unr) %>%
  summarise(
    human_cases=sum(human_c),
    human_pop=sum(human_pop),
    human_rate_mean=round(mean(human_cases_hab), 4),
    CI05 = round(quantile(human_cases_hab, probs = 0.05, na.rm = TRUE), 4),
    CI95 = round(quantile(human_cases_hab, probs = 0.95, na.rm = TRUE) , 4)) %>%
  mutate(level=cut(human_rate_mean, breaks = 3, labels = c("low","medium","high"))) %>% 
  arrange(disease, desc(human_rate_mean), UNregions)
  
datatable(lira_table_sp)
```

```{r}
lira_table_sp2 <- db %>%
  filter(time > 2003) %>%
  # filter(unr != "Western Asia") %>% 
  filter(!is.na(human_cases_hab)) %>%
  group_by(disease, UNregions=unr) %>%
  summarise(
    human_cases=sum(human_c),
    human_pop=sum(human_pop),
    human_rate_median=round(median(human_cases_hab), 4),
    CI05 = round(quantile(human_cases_hab, probs = 0.05, na.rm = TRUE), 4),
    CI95 = round(quantile(human_cases_hab, probs = 0.95, na.rm = TRUE) ,4)) %>%
  mutate(level=cut(human_rate_median, breaks = 3, labels = c("low","medium","high"))) %>% 
  arrange(disease, desc(human_rate_median), UNregions)

datatable(lira_table_sp2)
```
# Credits Acosta, Alfredo PhD<sup>1</sup>.
**SVA<sup>1</sup>**: SVA <http://www.sva.se/>. 